MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-05 rebuttal.txt

Q: Algorithm 1 should be made clearer: how are transition rates computed? What does the bit-flip mean? delta t is computed using a "u" - where does that come from?
A: Transition rates are computed by evaluating boolean formulas of each node, as is written on line 2 of Algorithm 1. To make it clearer, we modified the second paragraph of section 2.2, where we explicitly state, that the output of boolean logic is the transition rates. Regarding "u", the Gillespie algorithm uses 2 random numbers (r,u) drawn from the uniform distribution to randomly select the transition node and the transition time. We added comments to the algorithm and added a line referencing that the algorithm follows the Gillespie algorithm.

Q: Please ellaborate more on the last point in Figure 5, that seems to be an outlier. Which synthetic model is this? Is it also found in the earlier figures? 
A: Firstly, we included additional information about the synthetic model in the figure caption. Secondly, we analyzed the datapoint and it is the same case as in the previous MPI figure - the problem is distributed into so many small tasks that the actual cost of MPI communication becomes the bottleneck. We are sorry that we did not include this information in the first version - initially the figure was plotted in a different way which did not show this drastic speedup difference, after replotting the datapoint went unnoticed.

### 2024-05-05 Article.tex
explain outlier

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulae from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

@@

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where only a single node changes its value in each transition (contrary to the \emph{synchronous update strategy}, for which all nodes that can be updated are updated~\cite{stoll2012continuous}). Therefore, there is a transition from $S$ to $S'$ only if it holds that 
\begin{equation}
    \begin{split}
S_j \ne S'_j & \text{ for a given } j \\
S_i = S'_i & \text{ for } i \ne j.
    \end{split}
\end{equation}
Consequently, $S$ can have at most $n$ possible transitions. In programming terms, $S'$ is obtained by flipping the $j$-th bit of $S$.

@@

\begin{table}[h]
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 10\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{The main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of operands count in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/sizek_mpi.pdf}
\caption{Scalability results of MPI implementation on Sizek model with up to 192 MPI nodes and 20 cores per node, summing up to 3840 cores.}
\label{fig:sizek_results}
\end{figure}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/synth_mpi_speedup.pdf}
\caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and the formula size of 10, running on up to 192 MPI nodes with 32 cores per MPI node, summing up to 6144 cores.}
\label{fig:synthetic_results}
\end{figure}

To stress the scalability of the implementation, we also used a synthetic model with 1000 nodes. We simulated this model on 32 cores per MPI node, on 1 to 192 nodes (32 to 6144 cores). The obtained speedups are summarized in Figure~\ref{fig:synthetic_results}. Using this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As expected, the plateau in the speedup was not observed in simulations that involve larger models.

>>>>

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, database listings of such interactions~\cite{licata2020signor,turei2016omnipath} and experimental information retrieval techniques on a bigger number of components were subjected to many advancements. This resulted in the design of automatic methods for Boolean formulae inference from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

@@

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where only a single node changes its value in each transition (as opposed to the \emph{synchronous update strategy}, for which all nodes that can be updated are updated~\cite{stoll2012continuous}). Therefore, there is a transition from $S$ to $S'$ only if it holds that 
\begin{equation}
    \begin{split}
S_j \ne S'_j & \text{ for a given } j \\
S_i = S'_i & \text{ for } i \ne j.
    \end{split}
\end{equation}
Consequently, $S$ can have at most $n$ possible transitions. In programming terms, $S'$ is obtained by flipping the $j$-th bit of $S$.

@@

\begin{table}[h]
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 100\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{The main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of operands count in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/sizek_mpi.pdf}
\caption{Scalability results of MPI implementation on Sizek model simulating $1$ million trajectories with up to 192 MPI nodes and 20 cores per node, summing up to 3840 cores.}
\label{fig:sizek_results}
\end{figure}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/synth_mpi_speedup.pdf}
\caption{Speedup scaling of MPI implementation on the synthetic model with 1000 nodes, 100 million trajectories and the formula size of 10, running on up to 192 MPI nodes with 32 cores per MPI node, summing up to 6144 cores.}
\label{fig:synthetic_results}
\end{figure}

To stress the scalability of the implementation, we also used the synthetic model with 1000 nodes running $100$ million trajectories. We simulated this model on 32 cores per MPI node, on 1 to 192 nodes (32 to 6144 cores). The obtained speedups are summarized in Figure~\ref{fig:synthetic_results}. Using this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As expected, the plateau in the speedup was observed only for much bigger suites. More specifically, we can see a pronounced decrease in the speedup at $192$ nodes, hitting the aforementioned bottleneck during the utilization of more than $4096$ cores.
